---
title: "First Distill post"
description: |
  This is just a demo post. The idea is to test the capabilities of Distill as a blogging platform. In this post I am just trying out a few capabilities of Distill and how well it is suited to the kinds of things I would like to write about. Amongst other things. However, time will tell on how well I can focus on writing articles here.
date: 04-25-2021
bibliography: library.bib
output:
  distill::distill_article:
    self_contained: false
    toc: true
    toc_float: true
---

```{r setup, include=FALSE}
#knitr::opts_chunk$set(include = FALSE)
library(tidyverse)
library(knitr)
library(rmarkdown)
library(ggpubr)
library(hrbrthemes)
library(gtsummary)

trees <- read_delim("plot2.csv", ";", 
     escape_double = FALSE, locale = locale(decimal_mark = ",", 
         grouping_mark = "."), trim_ws = TRUE)
```

## Why Distill?

There are a few reasins for why I would like to try [Distill](https://rstudio.github.io/distill) as my blogging platform. As I want to write more things of scientific nature I would like to be able to easily cite articles. For example later in this example page I use volume functions from the article *Single-tree biomass and stem volume functions for eleven tree species used in Icelandic forestry* [@arnórsnorrason2006].

Other reasons are the capabilities of working with the data right in the blogging system and creating the charts and tables in the same area as I write the post.

## Working with data

For example we have the following data that I would like to work with.

```{r echo=FALSE}
paged_table(trees)
```

First I would like to do basic manipulation of the data.

```{r}
trees$dbh <- (trees$diam1 + trees$diam2) / 2
trees$ba = (trees$dbh/200)^2 * pi
```

For the next trick we use linear regressions to calculate the expected height of the trees. We then use `coalesce()` to copy back the height we already knew. From there we can calculate the volume of each tree using the volume functions [@arnórsnorrason2006].

```{r}
trees$height <- lm(height_measured ~ ba, data=trees) %>% 
  predict(trees)
trees$height <- coalesce(trees$height_measured, trees$height)

trees$v = 0.1299 * trees$dbh^1.6834 * trees$height^0.8598
```


```{r regression_plot, echo=FALSE, message=FALSE, warning=FALSE}
trees  %>% ggplot(aes(dbh, height_measured)) + 
  geom_point() +
  geom_smooth(method=lm) +
  stat_regline_equation() +
  stat_cor(label.y = 9, aes(label=..rr.label..)) +
  labs(x="DBH (cm)", y="Height (m)") +
  ggtitle("Regression of height") +
  theme_ipsum_rc()
```

## Math with R

Using LaTeX math symbols we can communicate mathematical functions in a nice manner. For example we can talk about BAL as described by Arne Pommerening's excellent article *[Basal area in larger trees and the growth compensation point](https://blogg.slu.se/forest-biometrics/2017/05/26/basal-area-in-larger-trees-and-the-growth-compensation-point/)* where he explains BAL as such:

> _BAL_ is related to available light, since with increasing basal area of larger trees there is less light available for smaller trees. In a sense _BAL_ is a surrogate for light measurements with the benefit that stem diameters and basal area are easier to measure.

$$BAL_i(t) = G(t) \cdot (1 - p_i(t)) \text{ where } p_i(t) = \frac{1}{G(t)} \sum_{\leq g_i(t)} g_i(t)$$

He also give us an example function in R. Let us use it to calculate the BAL of individual trees in our example data.

```{r}
bal <- function(ba, area) {
 sumba <- sum(ba)
 basmaller <- 0
 pix <- 0
 bal <- 0
 for (i in 1 : length(ba)) {
   bax <- ba[i]
   basmaller <- sum(ba[ba <= bax])
   pix <- basmaller / sumba
   bal[i] <- sumba * (1 - pix) / area
 }
 return(bal)
}
# An then run it for out 0.54ha example stand.
trees$bal <- bal(trees$ba, 0.54/10000)
```

